Fast Sorting on Modern Computers

### 2005-08-05 docs/tech_report/quick.tex
merged in david'd bit to quicksort. added a reference to mergresort. have yet to add replacement policy explanation

In the case that the array to be sorted fits in the cache, then quicksort has
excellent cache properties. This applies equally whether discussing a level 1
cache or a level 2 cache. Once the array is larger than the cache, however, the
number of misses increases dramatically. To fix this problem, LaMarca uses a
technique similar to that used in multi-mergesort. Instead of multi-merging a
series of sorted lists at the end, a \n{multi-way partition} is done at the
start, moving all keys into cache sized containers which can be sorted
individually.

@@

\subsection{Test Parameters}
%TESTING
The threshold chosen was 10, simply as this is the number used in Sedgewick's
book.

@@

LaMarca notes that quicksort is very cache efficient already. It has excellent
spatial and temporal locality. Once it no longer fits in the cache, however, the
number of misses should increase significantly. This is especially true in the
case of the base quicksort, which does an insertion sort across then entire
array at the end.

@@

The results show again that direct mapped caches can perform better than fully
associative ones. This is true in the case of all the quicksorts, as can be seen
in Figures \ref{base quicksort (no median) level 2 misses} to
\ref{multi-quicksort (sequential search) level 2 misses} graph that this is true
in the case of all the quicksorts.

@@

\section{Predictor Results}

%TODO a bit to rewrite
\subsection{Different Sorts}

\sixplot{base_quicksort__median_of_3__0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort__binary_search__0.eps}{multi_quicksort__sequential_search__0.eps}{}{}{base,
cache and multi quicksorts and base quicksort (pseudo-median-of-9)}

Firstly, memory-tuned quicksort and base quicksort (median-of-3) only differ in how they deal with
the insertion. Base quicksort (median-of-3) and base quicksort
(pseudo-median-of-9) have similar numbers of insertion branches, and a very
similar ratio (2/1 correct/incorrect) of mispredictions. Comparing base
quicksort to memory-quicksort (they both use
median of 3), there is actually a decrease in the number of branches handled by
insertion sort, instead of an increase. An increase is expected (thats why
sedgewick moves the insertion to the end), though when you take into account
the sentinel and the outer loop, there probably is an increase in instructions
at least. Base quicksort has 35\% miss rate, but cache has a 34\%. Not a big
difference, but the point is that there is a reduction in branches, but of the
reduced branches, 50\% were unpredictable. I dont think theres anything
interesting in this fact, however.

When going to multi-quicksort, there si quite a drop in the number fo branches.
Once the sort is larger than the cache, the number fo steps increases linearly
(since it does a sort of counting like radix sort). Theres a decrease of 19 \%.
Since the sort at this point is NlogM where M is the size of the cache (in
keys), there shouldnt be a change in misprediction rates. And there sint.
Before: 63\%. After, 63\%. So no difference.

Theres a lot of branches in the multi quicksort. WE saw in other results that
the mere instruction cost of this bogs down to algorithm significantly. in the
binary search version, there are more than twice as many mispredictions,
despite the fact that there are less than half the branches. (note that i dont
think this was a wonderful binary search, its heavily biased towards the
checking the left). binary search predicts correctly 57 percent of the time,
meaning that it mostly changes direction haphazardly during the search. No
wonder it performs so badly.

Note that sequential multi-quicksort has less mispredictions than memory-tuned quicksort.

%TODO to  here

\subsection{Different Medians}

@@

\label{quicksort entropy}

In Section \ref{heapsort entropy}, we observed that when an algorithm has more
branches than are strictly necessary complete, these extra branches are
predicted correctly. In quicksort, we expect to obtain similar results.

%HOW

We use several variations of quicksort to test this theory.
\cite{BentleyMcIlroy93} suggests partitioning using a median-of-9 instead of a
median-of-3. We created sorts using no median, a median-of-3, and
pseudo-medians-of-5, -7 and -9. As suggested, our pseudo-medians used repeated
medians-of-3 instead of a real median, in order to reduce the cost of their
calculation. It was also suggested that once the partition sized were small
enough, there was no more advantage to be gained by using a median of greater
than three, and so a median-of-3 was used below a threshhold of 40.

We measured the number of branches and mispredictions due to the partitioning,
the final insertion sort, and the partitioning. It is expected to have little or
no variance in either the number of branches and misses due to insertion sort,
but that less time will be spent partitioning to arrays as a better median is
found, and a corresponding increase in time spent finding the better median.

The results of these tests are shown in Figure \ref{Branch prediction
performance for base quicksort without a median, with median of 3, and with
pseudo-medians of 5, 7 and 9}. There is little variance in insertion sort misses
between the sorts using pseudo-medians, but the median-of-3 quicksort has
6\% fewer branches, and a corresponsing decrease in mispredictions.

Between a median-of-3 and a pseudo-median-of-9 partitioning, there is a
10\% reduction in branches. At the same time, there is a 20\% reduction in
correct predictions, accounting for the entire reduction in branches, and a 4\%
increase in mispredictions. This is due to better partitioning. Less time is
spent copying keys unnecessarily, resulting in a less predictable branch. The
corresponding increase in mispredictions while calculating the median equates to
50\% of the extra mispredictions due to partitioning.

This indicates the same pattern as we saw during heapsort.%
% and mergesort? TODO
The less entropy due to the partitioning key, the less wasted exchanging of keys,
which leads to a general decrease in branches. This decrease means that the
fewer keys are exchanged due to partitioning, leading to less predictable
partitioning. In effect, there are fewer branches pretending to be useful.

The pseudo-median-of-5 partitioning performs slightly better than the
median-of-3, and pseudo-median-of-7 performs slightly better again, in
terms of branch prediction. Figure \ref{quick_median_cycles} shows the cycle
count of each of the quicksorts, showing that a median-of-3 offers the greatest
trade-off between the number of instructions and the increase in branch
mispredictions.

\begin{figure}
\includegraphics{plots/quick_median_cycles.eps}
\caption{Empiric cycles count for quicksort with varying medians}
\label{quick_median_cycles}
\end{figure}

@@

The memory-tuned quicksort is the only algorithm here that requires a sentinel
check. The branch predictor will only mispredict this once, so the instruction
count increase is only that of removing a single check which occurs relatively
few ($N / THRESHOLD$) times\footnote{This is a minimum. The figure may be higher
since partitions are likely to be smaller than the threshold.}. Still, the
problem is interesting and has applications in other algorithm designs.

>>>>

In the case that the array to be sorted fits in a particular level of
the cache, then quicksort has excellent cache properties at that
level. Once the array is larger than the cache, however, the
number of misses increases dramatically. To fix this problem, LaMarca uses a
technique similar to that used in multi-mergesort. Instead of multi-merging a
series of sorted lists at the end, a \n{multi-way partition} is done at the
start, moving all keys into cache sized containers which can be sorted
individually.

@@

\subsection{Test Parameters}

When the size of a partition to be sorted is less than a certain
threshold value, insertion sort is used instead of quicksort. The
chosen threshold was 10, simply because this is the number used in
Sedgewick's book.

@@

LaMarca notes that quicksort is already very cache efficient as it has excellent
spatial and temporal locality. Once it no longer fits in the cache, however, the
number of misses should increase significantly. This is especially true in the
case of the base quicksort, which does an insertion sort across then entire
array at the end.

@@

The results show again that direct mapped caches can perform better than fully
associative ones. This is true in the case of all the quicksorts, as can be seen
in Figures \ref{base quicksort (no median) level 2 misses} to
\ref{multi-quicksort (sequential search) level 2 misses} graph that this is true
in the case of all the quicksorts. As in Section \ref{mergesort replacement
policy}, this is due to the replacement policy of the fully-associative cache.

@@

\section{Branch predictor results}

\subsection{More detailed analysis of branch prediction}

% \textbf{[David] I'm not sure what to do with this section. Part of the
% problem is that the pictures are very difficult to understand. Another
% problem is that we are trying to show differences that are almost
% invisible. The median of 9 picture should definitely be removed from
% this subsection. However, I suspect that the whole of this subsection should
% go. What do we want to say? One really puzzling result is that there
% seem to be no median branches at all for cache quicksort. This looks
% strikingly like a programming error.}

% \textbf{[David] On thing that's badly needed in this section is
% a detailed description of what these funny charts all mean. However,
% the appropriate place for most of this is the place where the first of
% these charts appeared.}

% \sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{cache_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}

\sixplot{base_quicksort__median_of_3__0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort__binary_search__0.eps}{multi_quicksort__sequential_search__0.eps}{}{}{base,
memory-tuned and multi quicksorts}

Figure \ref{base, cache and multi quicksorts}(a) shows the behaviour
of the indiviual branches in base quicksort. The $i$ branch is the
loop control branch for the index sweeping from left-to-right up the
array during partition. The $j$ branch is the corresponding loop
sweeping down the array. The partition step terminates when $i \geq
j$, and the $partition end$ branch checks that condition. The
$insertion$ branch is the comparison branch in the insertion sort
algorithm. Recall that our quicksort algorithm does not sort
partitions which are smaller than 10 elements, but instead does a
final pass of insertion sort over the whole array to sort these small
partitions. Finally, the $median$ branch is the sum of the results for
the branches involved in computing the median-of-three for each
partition.

The only difference between cache quicksort and base quicksort is that
cache quicksort applies insertion sort to small partitions
immediately, whereas base quicksort does a single insertion sort pass
at the end. The result is that both algorithms perform almost
identically from the viewpoint of branch prediction. Cache quicksort
uses a very slightly smaller number of insertion sort comparison
branches, because it sorts only within partitions and does not attempt
to move array elements that have acted as pivots. (The reason that
quicksort works is that after partitioning, the pivot is in its
correct location, and must never move again). In practice, this
reduction in branches is probably offset by the increase in
mispredictions in the outer loop exit branch of the many insertion
sorts.

Base quicksort has a 35\% branch misprediction rate, whereas the rate
for cache quicksort is only 34\%. The main reason for this is the
reduction in branches by not attempting to move the pivot elements.
Attempting to move these elements usually results in a branch
misprediction. The main reason is that most other elements move at
least one position, and thus cause the insertion sort loop to iterate
at least once. But the pivot elements never move --- they are already
in the correct place.

When we look at multi-quicksort, there is a 19\% drop in the number of
iterations of the $i$ and $j$ loops, with a corresponding reduction in
the number of mispredictions. Interestingly, the percentage
misprediction rate of the $i$ and $j$ branches does not change --- it
is 37\% in both cases.

% Once the sort is larger than the cache, the number fo steps increases linearly
% (since it does a sort of counting like radix sort). Theres a decrease of 19 \%.
% Since the sort at this point is NlogM where M is the size of the cache (in
% keys), there shouldnt be a change in misprediction rates. And there sint.
% Before: 63\%. After, 63\%. So no difference.

Multi-quicksort does, however, introduce a lot of additional
branches. We already saw in Figure \ref{quick instructions}, that the
binary search and sequential search variants increase the number of
executed instructions by around 25\% and 50\% respectively. Figure
\ref{base, cache and multi quicksorts} shows additional measures of the $binary
left$ and $binary right$ branches, which are the branches within the
binary search code. The number of binary search branches is much
greater than the reduction in $i$ and $j$ branches. Furthermore, these
binary search branches are highly (43\%) unpredictable.

The variation of multi-quicksort that uses sequential search dramatically
increases the number of executed branches, but these branches are highly
predictable. This version actually causes fewer branch mispredictions
than cache quicksort.

% Theres a lot of branches in the multi quicksort. WE saw in other results that
% the mere instruction cost of this bogs down to algorithm significantly. in the
% binary search version, there are more than twice as many mispredictions,
% despite the fact that there are less than half the branches. (note that i dont
% think this was a wonderful binary search, its heavily biased towards the
% checking the left). binary search predicts correctly 57 percent of the time,
% meaning that it mostly changes direction haphazardly during the search. No
% wonder it performs so badly.

% Note that sequential multi-quicksort has less mispredictions than cache quicksort.
\subsection{Different Medians}

@@

\label{quicksort entropy}
When we originally looked at quicksort, we expected that the $i$ and
$j$ branches in quicksort would resolve in each direction. In fact,
this is not the case. The main reason is that the chosen pivot is
almost never the true median of the section of the array to be
partitioned. Thus, the $i$ and $j$ branches are almost always biased,
which makes them predictable. According to Knuth, if the pivot element
is chosen randomly, then the $i$ and $j$ branches will be taken (i.e.
the $i$ or $j$ loop will execute) twice as often as they are not taken
(i.e. the loop exits).

In order to investigate this, we implemented a simple version of
quicksort that uses the middle element as the pivot rather than a
median-of-three\footnote{It is important to note that selecting
the pivot based on only a single element can be particularly inefficient
because it removes the natural sentinels at the top and bottom
of the regions to be partitioned that we get from median-of-three. In
Figure \ref{base quicksort without a median, with median of 3, and with pseudo medians of 5, 7 and 9} we see that this results in an additional branch to handle
the edgecase without a sentinel. Although this branch is almost 100\%
predictable, it is executed a very large number of times.}. The arrays we sort are random, so the value of this
element will also be random. Figure \ref{base quicksort without a median, with median of 3, and with pseudo medians of 5, 7 and 9}(a)
shows the behaviour of each of the important branches in this code. We
see that the $i$ and $j$ branches are indeed taken about twice as
often as not-taken, and the $i$ and $j$ branches are correctly
predicted around 71\% of the time.

The result that the branches are more predictable (71\%) than they are
biased (66\%) is a particularly interesting finding. The reason is that
the bias in the results is an average figure over a large number of
partitions that quicksort performs to sort the array. On a particular
partition, the bias will be more pronounced because the $i$ and $j$
branches will be biased in opposite directions.

For example, consider the case where the pivot element is larger than
the true median. In this case, the final resting place of the pivot
will be towards the right-hand end of the region to be
partitioned. Thus, the $i$ loop will execute a large number of times,
and thus the $i$ loop will be biased towards being taken. The $j$
loop, on the other hand, will be biased toward not-taken, because the
average element is likely to be smaller than the pivot. Therefore, the
$j$ loop will execute only a small number of times. Thus, we see that
in a particular partition, we do not normally get an equal number of
executions of the $i$ and $j$ loops. Instead we get a large number of
executions in one loop (which makes its branch relatively
predictable), whereas the other loop often does not enter at all
(which makes its branch also predictable, although executed a much
smaller number of times). Thus, although overall the $i$ and $j$
branches are taken twice as often as not-taken, on a particular
partition one will usually be biased in the not-taken direction, and
as a result will be executed a relativelty small number of times.

In the preceding paragraphs we have seen that the $i$ and $j$ branches
are predictable when we have a pivot that is larger or smaller than
the true median of the array section to be partitioned. An interesting
question is the effect of choosing a better approximation of the true
median on branch prediction. Figure \ref{base quicksort without a
median, with median of 3, and with pseudo medians of 5, 7 and 9}(b)
shows the behavior of these branches when median-of-three is used to
select the pivot. Perhaps the biggest effect is that the number of
executions of the $i$ and $j$ branches falls by around 15\%. This fall
is partly offset by the comparison branches to compute the median, but
these account for only around 1.25\% of the original branches from the
no-median quicksort. Thus, there is a reduction of almost 14\% in the
number of executed comparison branches. Even more interesting, however,
is the bias in the remaining $i$ and $j$ branches. The no-median $i$ and
$j$ branches are 66\% biased, but the median-of-three branches are
only 60\% biased in the taken direction. The accuracy with which
these branches are correctly predicted falls from 71\% to 64\%.

The reason for the reduction in bias and prediction accuracy is that
median-of-three gives a much better estimate of the true median of the
section to be partitioned. The final resting place of the pivot will
usually be closer to the middle of the section, to there is less bias
in the branches. Thus, although median-of-three gives around a 14\%
reduction in executed comparison branches, each of those branches is
much less likely to be correctly predicted. Overall there is actually
a 6\% \textit{increase} in the total number of branch mispredictions.

To further investigate this phenomenon, we measured the effects of
using pseudo-medians-of-five, -seven and -nine to choose the pivot. A
pseudo-median uses repeated medians-of-three for parts of the
computation to reduce the number of comparisons in computing the
pivot. The reductions in executed comparison branches are 16.5\%,
17.8\% and 19.6\% respectively. However, the biases of the $i$ and $j$
branches in these algorithms are only 57.8\%, 56.3\% and 55.1\%. The
corresponding predictabilities are 61.6\%, 59.9\% and 58.3\%. Overall,
the increases in comparison branch mispredictions (including computing
the pseudo-median) are 9.2\%, 11.8\% and 14.3\%.

Picking a better estimate of the true median substantially reduces the
number of executed branches, but each of those branches is much more
unpredictable. Intuitively, this makes sense. Sorting the array
involves putting an order on it, and thus reducing its entropy. If we
want to sort it in a smaller number of steps, then each step must
remove more randomness. Thus, we would expect that the outcome of the
branch that removes the randomness would itself be more random.

Figure \ref{TODO} shows the cycle count of each of the quicksorts,
showing that a pseduo-median-of-seven offers the greatest tradeoff
between the reduced number of branches and instructions, and the
increase in branch mispredictions.

An additional advantage of using a higher order median is that it
dramatically reduces the probability of quicksort exhibiting worst
case ($O(N^2)$) time complexity. Although the average case reduction
in running time is small, reducing the chance of worst-case
performance, which might cause the program to pause for minutes or
hours, is very valuable.

%TODO add a reference to this
\begin{figure}
\includegraphics{plots/quick_median_cycles.eps}
\caption{Empiric cycles count for quicksort with varying medians}
\label{quick_median_cycles}
\end{figure}

@@

Memory-tuned quicksort is the only algorithm here that requires a sentinel
check. The branch predictor will only mispredict this once, so the instruction
count increase is only that of removing a single check which occurs relatively
few ($N / THRESHOLD$) times\footnote{This is a minimum. The figure may be higher
since partitions are likely to be smaller than the threshold.}. Still, the
problem is interesting and has applications in other algorithm designs.
